The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-23 paper/paper-jsdse/paper-jsdse.tex
tidy data section

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative. Documenting the data cleaning is essential for the integrity of downstream statistical analyses and model building (M. Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis (IDA)'' (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis, which Tukey (1977) defines as detective work to learn from data, either numerically or graphically, before confirmatory data analysis is performed, which encompasses IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite its importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M. Huebner et al. 2020).

This paper demonstrates the steps of IDA -- tidying, cleaning, and documenting the process -- for a prominent open data source, National Longitudinal Survey of Youth (NLSY) 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but are not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, The NLSY is considered a survey with high retention rates and carefully designed, making it suitable for a life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousand of articles, and hundreds of book chapters and monographs has incorporated the NLSY data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the NLSY79 data sets (Pergamit et al. 2001).

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling. Our aim is to refresh this textbook data with data from . Here, we investigate the process of getting from the raw NLSY79 to this textbook data set. However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.

This paper is structured to have 5 sections. Section \ref{database} describes the original data source. Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and initial data analysis to find and treat the anomalies in the data. We also gave examples of exploratory data analysis using the clean data in Section \ref{eda}. Finally, Section \ref{summary} summarises the paper.

@@

\hypertarget{target-data}{%
\subsection{Target Data}\label{target-data}}

As mentioned in Section \ref{intro}, this paper aims to refresh the NLSY79 data as used in Singer and Willet's textbook (Singer and Willett (2003)). This data set contains the information of yearly mean hourly wages from the NLSY79 cohort, with education and race as covariates. This data set represents the measurement of male high-school dropouts, aged from 14 to 17 years old when first time measured. Accordingly, to get this data, we have two additional target data, NLSY79 cohort's demographic profiles and wages data of people who completed education up to high school.

Since there is no information on high school dropouts in Singer and Willet's book, we define high school dropouts as respondents who only completed either \(9^{th}\), \(10^{th}\), \(11^{th}\) grade, or who completed \(12^{th}\) when their age is at least 19 years old (means that these IDs being dropped out in a certain year, but they returned back to high school to complete it).

@@

The NLSY79 data are stored in a \href{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}{database} and could be downloaded by variables. Several variables are available for download and could be browsed by index. For the wages data set, we only extracted these variables:

@@

According to Wickham (2014), tidy data sets comply with three rules. Firstly, each variable forms a column. Secondly, each observation forms a row. Lastly, each type of observational unit forms a table. Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) to do this job.

>>>>

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (M. Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis (IDA)'' (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M. Huebner et al. 2020).

This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but are not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling. Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).

Our process of cleaning builds heavily on the tidyverse approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

This paper is structured in the following way. Section \ref{database} describes the NLSY79 data source. Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies. Our final subset is compared the old textbook subset in Section \ref{compare}. Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.

@@

\hypertarget{target}{%
\subsection{Target Data}\label{target}}

This paper aims, as mentioned in Section \ref{intro}, to refresh the NLSY79 data used in Singer and Willett (2003). This data set contains the information of yearly mean hourly wages from the NLSY79 subcohort, with education and race as covariates. The subcohort correspond to male high-school dropouts that participated in the study from 14 to 17 years old. Accordingly, to refresh this data, we make use of NLSY79 cohort's demographic profiles and wages data of people who completed education up to high school.

Since no specific criteria on high school dropouts was defined
in Singer and Willett (2003), we define high school dropouts as respondents who only completed either 9th to 11th grade, or who completed 12th when their age is at least 19 years old (meaning that the individual may have dropped out of high school but returned at a later time to complete high school education).

@@

The NLSY79 data contains a large scope of variables but we limit the scope of the data to demographic profiles and wages data. To extract the relevant component of the data, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} and navigated as described below.

@@

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The input data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1980}, such as contain the information about the job number and the year. The input data consequently has a large number of columns ()

Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) to do this job.

### 2021-09-23 paper/paper.Rmd
tidy data section

The NLSY79 data are stored in a [database](https://www.nlsinfo.org/content/cohorts/nlsy79/get-data) and could be downloaded by variables. Several variables are available for download and could be browsed by index. For the wages data set, we only extracted these variables:

::: {.tcolorbox data-latex="[label=\"box:data\", title=Getting the data]" }
\faDatabase\ National Longitudinal Survey of Youth 1979

\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCloudDownload] Downloaded all of variables in Highest grade completed.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job $\triangleright$ Hours worked and Hourly wages
\begin{itemize}
\item[\faCloudDownload] Downloaded all of the variables in Hours worked.
\item[\faCloudDownload] Downloaded all of the variables in Hourly wages
  Both hours worked and hourly wages are recorded by the job, up to five jobs for each id/subject. 
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics 
\begin{itemize}
\item[\faCloudDownload] Downloaded year and month of birth, race, and sex variables. There are two versions of the year and month of birth, i.e., 1979 and 1981 data. We downloaded these two versions.
\end{itemize}
\end{itemize}
\end{itemize}

@@

According to @WickhamHadley2014TD, tidy data sets comply with three rules. Firstly, each variable forms a column. Secondly, each observation forms a row. Lastly, each type of observational unit forms a table. Unfortunately, the `new_data_qnames` did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used `tidyr` [@tidyr], `dplyr` [@dplyr], and `stringr` [@stringr] to do this job.

>>>>

The NLSY79 data contains a large scope of variables but we limit the scope of the data to demographic profiles and wages data. To extract the relevant component of the data, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data} and navigated as described below.

::: {.tcolorbox data-latex="[label=\"box:data\", title=Getting the data]" }
\faDatabase\ National Longitudinal Survey of Youth 1979 
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCloudDownload] Downloaded all of variables in Highest grade completed.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job $\triangleright$ Hours worked and Hourly wages
\begin{itemize}
\item[\faCloudDownload] Downloaded all of the variables in Hours worked.
\item[\faCloudDownload] Downloaded all of the variables in Hourly wages
  Both hours worked and hourly wages are recorded by the job, up to five jobs for each id/subject. 
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics 
\begin{itemize}
\item[\faCloudDownload] Downloaded year and month of birth, race, and sex variables. There are two versions of the year and month of birth, i.e., 1979 and 1981 data. We downloaded these two versions.
\end{itemize}
\end{itemize}
\end{itemize}

@@

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The input data, `new_data_qnames`, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1980`, such as contain the information about the job number and the year. The input data consequently has a large number of columns (686 to be specific).

[REMOVE] Unfortunately, the `new_data_qnames` did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used `tidyr` [@tidyr], `dplyr` [@dplyr], and `stringr` [@stringr] to do this job.

### 2021-09-23 paper/paper.tex
tidy data section

\hypertarget{target-data}{%
\subsection{Target Data}\label{target-data}}

@@

According to Wickham (2014), tidy data sets comply with three rules. Firstly, each variable forms a column. Secondly, each observation forms a row. Lastly, each type of observational unit forms a table. Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) to do this job.

>>>>

\hypertarget{target}{%
\subsection{Target Data}\label{target}}

@@

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The input data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1980}, such as contain the information about the job number and the year. The input data consequently has a large number of columns (686 to be specific).

{[}REMOVE{]} Unfortunately, the \texttt{new\_data\_qnames} did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019) to do this job.
